### 2024-05-26 getting-started.qmd
20240526 - join types

>>>>

## Merging/Joins {#sec-merging}

### Overview {#sec-mergingOverview}

Merging (also called joining) merges two data objects using a shared set of variables called "keys."
The keys are the variable(s) that uniquely identify each row (i.e., they account for the levels of nesting).
In some data objects, the key might be the participant's ID (e.g., `participantID`).
However, some data objects have multiple keys.
For instance, in long form data objects, each participant may have multiple rows corresponding to multiple timepoints.
In this case, the keys are `participantID` and `timepoint`.
If a participant has multiple rows corresponding to timepoints and measures, the keys are `participantID`, `timepoint`, and `measure`.
In general, each row should have a value on each of the keys; there should be no missingness in the keys.

To merge two objects, the keys must be present in both objects.
The keys are used to merge the variables in object 1 (`x`) with the variables in object 2 (`y`).
Different merge types select different rows to merge.

Note: if the two objects include variables with the same name (apart from the keys), `R` will not know how you want each to appear in the merged object.
So, it will add a suffix (e.g., `.x`, `.y`) to each common variable to indicate which object (i.e., object `x` or object `y`) the variable came from, where object `x` is the first object—i.e., the object to which object `y` (the second object) is merged.
In general, apart from the keys, you should not include variables with the same name in two objects to be merged.
To prevent this, either remove or rename the shared variable in one of the objects, or include the shared variable as a key.
However, as described above, you should include it as a key ***only*** if it uniquely identifies each row in terms of levels of nesting.

### Data Before Merging {#sec-mergingData}

Here are the data in the `nfl_actualStats_offense_weekly` object:

```{r}
nfl_actualStats_offense_weekly

dim(nfl_actualStats_offense_weekly)

dim(distinct(nfl_actualStats_offense_weekly, player_id, season, week, .keep_all = TRUE))

str(nfl_actualStats_offense_weekly[,c("player_id", "season", "week")])
```

The data are structured in player-season-week form.
That is, every row in the dataset is uniquely identified by the variables, `player_id`, `season`, and `week`.

Here are the data in the `nfl_expectedFantasyPoints_weekly` object:

```{r}
nfl_players

dim(nfl_players)

dim(distinct(nfl_players, gsis_id, .keep_all = TRUE))
```

### Types of Joins {#sec-mergeTypes}

#### Visual Overview of Join Types {#sec-mergeTypesVisual}

Below is a visual that depicts various types of merges/joins.
Object `x` is the circle labeled as `x`.
Object `y` is the circle labeled as `y`.
The area of overlap in the Venn diagram indicates the rows on the keys that are shared between the two objects (e.g., the same `player_id`, `season`, and `week`).
The non-overlapping area indicates the rows on the keys that are unique to each object.
The shaded blue area indicates which rows (on the keys) are kept in the merged object from each of the two objects, when using each of the merge types.
For instance, a left outer join keeps the shared rows and the rows that are unique to object `x`, but it drops the rows that are unique to object `y`.

![Types of merges/joins](images/joinTypes.png)

#### Full Outer Join {#sec-fullJoin}

A full outer join includes all rows in $x$ **or** $y$.
It returns columns from $x$ and $y$.
Here is how to merge two data frames using a full outer join (i.e., "full join"):

```{r}
fullJoinData <- full_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

fullJoinData
dim(fullJoinData)
```

#### Left Outer Join {#sec-leftJoin}

A left outer join includes all rows in $x$.
It returns columns from $x$ and $y$.
Here is how to merge two data frames using a left outer join ("left join"):

```{r}
leftJoinData <- left_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

leftJoinData
dim(leftJoinData)
```

#### Right Outer Join {#sec-rightJoin}

A right outer join includes all rows in $y$.
It returns columns from $x$ and $y$.
Here is how to merge two data frames using a right outer join ("right join"):

```{r}
rightJoinData <- right_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

rightJoinData
dim(rightJoinData)
```

#### Inner Join {#sec-innerJoin}

An inner join includes all rows that are in **both** $x$ **and** $y$.
An inner join will return one row of $x$ for each matching row of $y$, and can duplicate values of records on either side (left or right) if $x$ and $y$ have more than one matching record.
It returns columns from $x$ and $y$.
Here is how to merge two data frames using an inner join:

```{r}
innerJoinData <- inner_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

innerJoinData
dim(innerJoinData)
```

#### Semi Join {#sec-semiJoin}

A semi join is a filter.
A left semi join returns all rows from $x$ **with** a match in $y$.
That is, it filters out records from $x$ that are not in $y$.
Unlike an [inner join](#innerJoin), a left semi join will never duplicate rows of $x$, and it includes columns from only $x$ (not from $y$). 
Here is how to merge two data frames using a left semi join:

```{r}
semiJoinData <- semi_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

semiJoinData
dim(semiJoinData)
```

#### Anti Join {#sec-antiJoin}

An anti join is a filter.
A left anti join returns all rows from $x$ **without** a match in $y$.
That is, it filters out records from $x$ that are in $y$.
It returns columns from only $x$ (not from $y$).
Here is how to merge two data frames using a left anti join:

```{r}
antiJoinData <- anti_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

antiJoinData
dim(antiJoinData)
```

#### Cross Join {#sec-crossJoin}

A cross join combines each row in $x$ with each row in $y$.

```{r}
crossJoinData <- cross_join(
  unique(nfl_actualStats_offense_weekly[,c("player_id","player_display_name")]),
  data.frame(season = 2020:2024))

crossJoinData
dim(crossJoinData)
```
